Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]

### 2025-12-09 paper/manuscript.tex
Minor tweaks

The chimeric constructs needed in single and dual topology approaches in the sense of Pearlman’s definition just described quickly become non-trivial to set up and require care to avoid artifacts (see below). %\sbnote{See below; where do we place the dummy atom discussion?}.
Some of these difficulties might be avoided by having the initial and final state present as separate molecules with separate coordinate sets. To the best of our knowledge, such an approach was first described by Axelsen and Li \cite{Axelsen_1998}. Both initial and final state are present simultaneously; the two copies never see each other and interactions with the environment are turned on/off as a function of $\lambda$. Harmonic restraints between corresponding atoms in the initial and final state copy prevent the two molecules (sets of atoms from drifting apart). %Unfortunately, Axelsen and Li referred to their ansatz also as dual-topology, though in practice it is quite different from Pearlman’s meaning of the term.
A recent implementation of this idea is Restraintmaker \cite{Ries_2022}, a tool to automatically assign the distance restraints needed to keep the two molecules / regions (one for each of the physical endstates) on top of each other. The idea is illustrated in the right branch of Fig.~\ref{fig:fig_topology}.  Somewhat unfortunately, Axelsen and Li, as well as Ries et al., refer to their approach as dual topology. Ries et al suggested to use the term hybrid topology for (single coordinate) dual topology in the sense of Pearlman. We will attempt a disambiguation of the various nomenclatures below.

@@

It is difficult to make general statements about the behavior of $\rho_t (x | x_0)$. For systems not involving a protein, it may be possible to identify clear "plateaus'' of convergence \footnote{The relevant mathematical object here is a quasi-stationary distribution, but should we say this? Could cite Art Voter at Los Alamos, maybe even Freidlin-Wentzell for the original quasipotential.}, i.e. intervals $[T_a, T_b]$, $T_a \ll T_b$, within which $\rho_t (x | x_0)$ does not vary much for $t \in [T_a, T_b]$, and is roughly invariant to $x_0 \in B$ in some appropriate "starting domain". Consider, for instance, the estimation of free energies of hydration for small molecules [cite Shirts paper on free energies of hydration]~\cite{paliwal2011benchmark}. In this case, the free energy estimates behave in accordance with classical statistical theory, and the density $\rho_t (x | x_0)$ should not vary much along coordinates relevant to the free energy estimates.

@@

%% MS: I'm not tied to any particular definition of ensemble, but we should probably make a concrete one:
% - common x_0 ? 
% - v ~ MaxwellBoltzmann(Temperature) 
% - what are the individual trajectories called? "replicates" ? 
We define an \textbf{ensemble} to be a set $\{ X^{(1)}, \ldots, X^{(R)} \}$ of $R \geqslant 1$ trajectories, all started at the same configuration $x_0$, with $v_0$ drawn from the Maxwell-Boltzmann distribution at the appropriate temperature, and run for a chemical time $T$.

\subsubsection{Ensemble error estimates}

The simplest  and most robust way of obtaining error bars on free energy estimates is by simulating an ensemble $\{ X^{(1)}, \ldots, X^{(R)} \}$ of $R$ molecular dynamics trajectories started at $x_0$ and producing trajectories of free energy differences $\{ \Delta G^{(1)}, \ldots, \Delta G^{(R)} \}$. The trajectories $\Delta
G^{(i)} (\cdot)$ are identically distributed and conditionally independent given some $x_0$. For time $t > 0$, the variance of any $\Delta G^{(i)} (t)$ can be estimated with the sample variance $S^2_{(R - 1)} (t)$,
\begin{equation}
  S^2_R (t) = \frac{1}{R - 1} \sum_{i = 1}^R (\Delta G^{(i)} (t) -
  \overline{\Delta G (t)})^2, \qquad \overline{\Delta G (t)} = \frac{1}{R}
  \sum_{i = 1}^R \Delta G^{(i)} (t) .
\end{equation}
Note that, conditionally on $x_0$, we have
\begin{equation}
    \mathbb{E}_{x_0}[S_R^2(t)] = \mathrm{Var}_{x_0}(\Delta G^{(1)}(t))
\end{equation}
The above equation says that $S_R^2(t)$ is an unbiased estimator for the variance. By a law of large numbers, as $R \rightarrow \infty$, we have $S^2_R (t) \rightarrow \mathrm{Var} _{x_0}(\Delta
G^{(1)} (t)) = \sigma^2 (t ; x_0)$. 
%% Should we say what kind of LLN (weak/strong) ? Do we provide a reference (standard: Durrett or Chung)?

@@

Alchemical Free Energy (AFE) calculations have primarily been applied to simulate the stoichiometric, reversible binding of organic molecules to well-structured proteins. However, several areas of drug discovery involve other classes of binding events. While the underlying theory of alchemical free energy calculations is generally applicable, additional work may be required to develop efficient protocols suited to systems beyond the traditional small molecule–protein complex.
For example, there is growing interest in identifying small molecules that directly bind to RNA or DNA, with recent preliminary studies showing promising results using specialized simulation protocols.\cite{Abramyan2025,Rasouli2024} In contrast, there has been limited investigation into the applicability of AFE methods to small-molecule ligands targeting intrinsically disordered proteins (IDPs,\cite{Papadourakis2024} or proteins containing intrinsically disordered regions (IDRs) that undergo disorder-to-order transitions upon ligand binding.\cite{MendozaMartinez2022}
The treatment of covalent binders has also received some attention.\cite{luo2021,zhang2019} However, robust protocols for handling datasets of irreversible covalent binders containing diverse ‘warheads’ have not yet achieved widespread adoption.\cite{Yu2019} A variety of other applications are less traditional and may be possible but raise new practical issues; for example, if multiple copies of a ligand bind at once in the same site, this could pose challenges, as could handling of binding and function prediction for molecular glues and related categories of molecules.

>>>>

The chimeric constructs needed in single and dual topology approaches in the sense of Pearlman’s definition just described quickly become non-trivial to set up and require care to avoid artifacts (see below). %\sbnote{See below; where do we place the dummy atom discussion?}.
Some of these difficulties might be avoided by having the initial and final state present as separate molecules with separate coordinate sets. To the best of our knowledge, such an approach was first described by Axelsen and Li \cite{Axelsen_1998}. Both initial and final state are present simultaneously; the two copies never see each other and interactions with the environment are turned on/off as a function of $\lambda$. Harmonic restraints between corresponding atoms in the initial and final state copy prevent the two molecules (sets of atoms from drifting apart). %Unfortunately, Axelsen and Li referred to their ansatz also as dual-topology, though in practice it is quite different from Pearlman’s meaning of the term.
A recent implementation of this idea is Restraintmaker \cite{Ries_2022}, a tool to automatically assign the distance restraints needed to keep the two molecules / regions (one for each of the physical endstates) on top of each other. The idea is illustrated in the right branch of Fig.~\ref{fig:fig_topology}.  Somewhat unfortunately, Axelsen and Li, as well as Ries et al refer to their approach as dual topology. Ries et al suggested to use the term hybrid topology for (single coordinate) dual topology in the sense of Pearlman. We will attempt a disambiguation of the various nomenclatures below.

@@

It is difficult to make general statements about the behavior of $\rho_t (x | x_0)$. For systems not involving a protein, it may be possible to identify clear "plateaus'' of convergence \footnote{The relevant mathematical object here is a quasi-stationary distribution, but should we say this? Could cite Art Voter at Los Alamos, maybe even Freidlin-Wentzell for the original quasipotential.}, i.e. intervals $[T_a, T_b]$, $T_a \ll T_b$, within which $\rho_t (x | x_0)$ does not vary much for $t \in [T_a, T_b]$, and is roughly invariant to $x_0 \in B$ in some appropriate "starting domain". Consider, for instance, the estimation of free energies of hydration for small molecules [cite Shirts paper on free energies of hydration]. In this case, the free energy estimates behave in accordance with classical statistical theory, and the density $\rho_t (x | x_0)$ should not vary much along coordinates relevant to the free energy estimates.

@@

%% MS: I'm not tied to any particular definition of ensemble, but we should probably make a concrete one:
% - common x_0 ? 
% - v ~ MaxwellBoltzmann(Temperature) 
% - what are the individual trajectories called? "replicates" ? 
We define an \textbf{ensemble} to be a set of $R \geqslant 1$ trajectories, all started at the same configuration $x_0$, with $v_0$ drawn from the Maxwell-Boltzmann distribution at the appropriate temperature, and run for a chemical time $T$.

\subsubsection{Ensemble error estimates}
The simplest  and most robust way of obtaining error bars on free energy estimates is by simulating an ensemble $\{ X^{(1)}, \ldots, X^{(R)} \}$ of $R$ molecular dynamics trajectories started at $x_0$ and producing trajectories of free energy differences $\{ \Delta G^{(1)}, \ldots, \Delta G^{(R)} \}$. The trajectories $\Delta
G^{(i)} (\cdot)$ are identically distributed and conditionally independent given some $x_0$. For time $t > 0$, the variance of any $\Delta G^{(i)} (t)$ can be estimated with the sample variance $S^2_{(R - 1)} (t)$,
\begin{equation}
  S^2_R (t) = \frac{1}{R - 1} \sum_{i = 1}^R (\Delta G^{(i)} (t) -
  \overline{\Delta G (t)})^2, \qquad \overline{\Delta G (t)} = \frac{1}{R}
  \sum_{i = 1}^R \Delta G^{(i)} (t) .
\end{equation}
Note that, conditionally on $x_0$, we have
\begin{equation}
    \mathbb{E}_{x_0}[S_R^2(t)] = \mathrm{Var}_{x_0}(\Delta G^{(1)}(t))
\end{equation}
The above equation says that $S_R^2(t)$ is an unbiased estimator for the variance. By a law of large numbers, as $R \rightarrow \infty$, we have $S^2_R (t) \rightarrow \mathrm{Var} _{x_0}(\Delta
G^{(1)} (t)) = \sigma^2 (t ; x_0)$. 
%% Should we say what kind of LLN (weak/strong) ? Do we provide a reference (standard: Durrett or Chung)?

@@

Alchemical Free Energy (AFE) calculations have primarily been applied to simulate the stoichiometric, non-reversible binding of organic molecules to well-structured proteins. However, several areas of drug discovery involve other classes of binding events. While the underlying theory of alchemical free energy calculations is generally applicable, additional work may be required to develop efficient protocols suited to systems beyond the traditional small molecule–protein complex.
For example, there is growing interest in identifying small molecules that directly bind to RNA or DNA, with recent preliminary studies showing promising results using specialized simulation protocols.\cite{Abramyan2025,Rasouli2024} In contrast, there has been limited investigation into the applicability of AFE methods to small-molecule ligands targeting intrinsically disordered proteins (IDPs,\cite{Papadourakis2024} or proteins containing intrinsically disordered regions (IDRs) that undergo disorder-to-order transitions upon ligand binding.\cite{MendozaMartinez2022}
The treatment of covalent binders has also received some attention.\cite{luo2021,zhang2019} However, robust protocols for handling datasets of irreversible covalent binders containing diverse ‘warheads’ have not yet achieved widespread adoption.\cite{Yu2019} A variety of other applications are less traditional and may be possible but raise new practical issues; for example, if multiple copies of a ligand bind at once in the same site, this could pose challenges, as could handling of binding and function prediction for molecular glues and related categories of molecules.
